Dynamics of structure (DoS) is an open source, free web based software, developed by students of 
Testing and Consultancy Cell (TCC), under guidance of Dr. HS Rai. This software is used to compute the 
Modes of vibration in which structure can move and also force applied on each floor 
due to the vibration caused by the earthquake. so, that civil engineers can
analysis the stability of structure consisting of many stories.

 The main task of this software is to get data as input from user and then it can compute 
the result at back end and when the result is ready it send output as email.
This software is structured keep in view that user of this software can be both 
a civil engineer or a simple man whose job is just to enter data to software 
in order that an engineer can analysis latter from result stability of the structure.
This software also provide the intermediate values for engineering student 
to deeply analysis the process of computation to be done in order to get output values. 


The core part of DoS is implemented using Sagemath for processing,  \LaTeX{} for output generation and django 
for web interface. shell script is used to join all these and bootstrap to make 
the web application responsive.


My being not based on particular language or technology different type of open softwares and technologies are 
used in this project and many during my training which are not used in this 
project like OpenCV (for image processing and computer vision), CGI (for web interface through c++).
