Dynamics of Structures (DoS) is an open-source, free web-based software, developed by students of 
Testing and Consultancy Cell (TCC), under the guidance of Dr. H.S. Rai. This software is used to compute the 
Modes of vibration in which the structure can move and also force applied on each floor 
due to the vibration caused by the earthquake. So, that Civil engineers can
analysis the stability of structure consisting of many stories.

 The main task of this application is to get data as input from user and then it can compute 
the result at back-end and when the result is ready it send output as email.
This software is structured by keeping in view that user of this software can be both 
a Civil Engineer or a simple man whose job is just to enter data to software 
in order that an engineer can analyze later from the result stability of the structure.
This software also provides the intermediate values for engineering student 
to deeply analyze the process of computation to be done in order to get output values.


The core part of DoS is implemented using Sagemath for processing,  \LaTeX{} for output file (PDF) generation, django 
for web interface. Bash Shell Scripting has been used to join all these. To provide the User Experience to the users, 
Bootstrap has been used.


My training being not based on particular language or technology, different type of open-source softwares and technologies are 
used in this project and many during my training which are not used in this 
project like OpenCV (for image processing and computer vision), CGI (for web interface through c++).
